\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{makeidx}

\title{Minor Culture and Cognition\\Notes and Summary}
\author{Rowan Rodrik van der Molen}

\begin{document}

\maketitle

\tableofcontents

\section{Evolution of the humanities}

The humanities as a field of study started in the 19th century, in response to the rise of nation states and atheism, as a new style of group identity.
A recurring argument from the humanities is that culture and humanity are somehow special, apart from nature and thus not subject to study by the natural sciences. The humanities vs. evolutions debate smells of creationism.

The humanities are losing money; they're having trouble legitimizing themselves as science proper while the naturalists are encroaching in their territory. Therefore, there is increasing interest in multidisciplinarity. Hence the research group that organizes this minor, which bridges the gap between ‘nature’ and ‘culture’---between brain matter and mind.



\citet{mesoudi2006}


\section{An interdisciplinary approach to the evolution of culture and cognition}

Biology, neuroscience, psychology, ethology, anthropology

Behaviourism (mind as a black box, studying inputs and outputs) was followed by cognitive science in 1957. Later came Noam Chomsky with his Universal Grammar / language device. His were the beginnings of theory, but not yet evolutionary theory.

Darwin's thesis: 4 requirements for evolution: variation, selection, heridity, continuity.

\citet{donald1991} offers a \textbf{gradualist} hypothesis, wherein the language faculty is not a discontinuos trait as it is in Noam Chomsky's \textbf{Universal Grammar} Hypothesis. The latter is a \textbf{saltationist} (\textit{salto}=big jump) approach to language. Saltationism, like \textbf{creationism}, is a \textbf{discontinuity} approach, and, like creationism, it violates one of the basic requirements for an evolutionary theory: \textbf{continuity}, which Donald's hypothesis does support.
\index{Universal Grammar}
\index{continuity}
\index{gradualism}
\index{saltationism}
\index{creationism}
% I'm not sure if my above statements are really accurate. Chomsky doesn't strike me as the discontinuous type.

Corballis: generativity $\ne$ discontinuity. Generativity: limited set of rules --> Unlimited set of phenomena.

\section{Levels of cognition}

Memory is a set of representations. “The human mind evolved from the primate (episodic) mind through a series of major adaptations, each of which lef to the emergence of a new representational system” \citep{donald1991}. The translation of memory to behaviour has become increasingly complex with each new level of representation.

\citet{donald1991} describes 4 stages of ape/hominid culture, each of which represents a new level of cultural complexity that emerges at the critical threshold of a gradual increase in cognitive abilities:

\begin{enumerate}
  \item episodic culture (non-semiotic), also present in the other extant great apes, emerged from the ability to represent and relate to events in memory;
  \item mimetic culture, allowed by iconic semiosis, emerged from the ability to model the self such that current behaviour can be linked to past, current or future events (metacognition);
  \item mythic culture emerged from the ability to invent semiotic devices (mind-tools/symbols); and
  \item theoretic culture emerged from our argumentative ability.
\end{enumerate}

Each new level of cultural complexity has depended on the evolution of new . “Each successive new representational system has remained intact within our current mental architecture, so that the modern mind is a mosaic structure of cognitive vestiges from earlier stages of human emergence.” \citep{donald1991}

\citet{collins2013} has a simpler take: System 1 ---> System 2

\index{semiosis}
Semiosis = the use of signs, the basis of communication.

\subsection{Episodic culture}


\subsection{Mimetic culture}

Aristotle on tragedies: Imitation is good
Katharsis requires a certain distance between the work of art/artist and the viewer as well as identification. Memesis allows us to empathize. [According to Miranda, that is. I know for a fact that chimps can empathize, and their culture is episodic.]

An example of mimesis: using a water bottle to signify a spear.

Art is a type of cognitive engineering tool to manipulate the minds of others.

A poet plays with the space between literal re-enactment and convention by turning words into objects---the process of femaliarization/enstrangement (Heidigger).

\subsubsection{Group selection}

Donald argues that, with the advent of mimetic culture, selection no longer operated solely at the level of the individual, because group selection came into play.

\subsection{Mythic culture}

Apes can be trained in symbol use, but they don't invent symbols. This requires more than a capacity for linking signs and signifiers \citep[p.~217-220]{donald1991}. A symbol can be considered a mind-tool \citep{gregory1981} and ``must have followed an advance in thought skills'' \citep[p.219]{donald1991}.

Mimetics: perceptual mental models

Symbols: conceptual mental models

Causes of mimetic ---> mythic transition:
\begin{itemize}
  \item Speech adaptation
  \item Not geographic or climatic conditions
  \item Competition between groups of hominids instead of individuals % ESSAY
  \item Primary human adaptation was mythical thinking, not language. Symbols were needed for these new mental models.
\end{itemize}

\subsubsection{Structuralism}

Saussure (1857-1913).

signifi\'{e} = phonetic aspect of the sign
significant = semantic concept
Binary oppositions: chair $\ne$ table

\subsubsection{Linguistic turn}

Mid 20th century.
Language as focal point  in understanding historical representation.
Subjectivity
Gender. Women were no longer considered merely as a peculiar type of men.

\textbf{Tropes} connect concepts in different ways and have a function in group formation, by seperating the hearers of message 1 and message 2.

\begin{enumerate}
  \item \textbf{Metaphor}
  \item \textbf{Metonymy}
  \item \textbf{Synecdote}
  \item \textbf{Irony}, the basis of humor, consists of a message that is clearly acted out and a (subliminal) message that is represented but not acted out. The subliminal message could be hidden in the tone of voice, posture, or in the mimetic representation of the event to contrast with the symbolic content of the verbal report.
\end{enumerate}

Meaning is created by \textbf{discourse}---“practices that systematically form the objects of which they speak (Foucault, 1926-1984).” Implications:
\begin{itemize}
  \item Anti-humanism (I only exist as a crossroads)
  \item Knowledge/power relation
  \item Linguistic being trapped in symbol space
\end{itemize}

\textbf{Deconstruction} (Derrida, 1930-2004): Unpack hierarchy of meaning implicit within dichotomous thinking.

\subsection{Theoretic culture}

\section{Speech}

Prosody

Phonology
Parralel with prosody system \citep[p.~249]{donald1991}.

In pure Wernicke's aphasia, phonology and morphology of language remains intact, whereas semantics and syntax suffer. \citep[p.~249]{donald1991}.

lexicons (vs dictionaries) \citep[fig.~7.1, p.~251]{donald1991}.

Two major modes of thought: narrative and paradigmatic.

The myth as the supreme product of the narrative mode.

\section{Brain structures}

\paragraph{Aphasics}

\paragraph{Wernicke's region}

Different effects of damage to this region \citep[p.~262]{donald1991}: agrammatism and reading+writing.

Phonology is highly laterized to the left hemisphere \citep[p.~262]{donald1991}.

\section{Biological (brain) evolution}
\label{sec:biological_evolution}

\textbf{The default hypothesis} is the hypothesis that typical human (system 2) features
of the brain evolved in the left side of the brain (Korbalis).

\begin{tabular}{rl}
5 mya & common ancestor homonids \& chimps \\
4 mya & \textit{Australopithecus} (“Southern ape”) \\
2 mya & \textit{Homo habilis} (“Handy man”) \\
1.5 mya & \textit{Homo erectus} \\
0.3 mya & (Archaic) \textit{Homo sapiens} \\
0.04 mya & \text{H. sapiens sapiens} \\
\end{tabular}

\paragraph{Hominidae}

\paragraph{\textit{Homo habilis}} lived 2.3 to 1.6 mya and created the first manufactored tools (`Oldowan' tools).

\paragraph{\textit{Homo erectus}} lived 1.9 mya to 45 kya.
Oldowan and Acheuléen tools.

The development of an advanced vocal apparatus would have followed from the selection pressure from a mind that needed vocal language for its modeling \citep[p.220]{donald1991}.

\subsection{Four-dimensional evolution}

\citet{jablonka2007}


\section{Cultural evolution}
\label{sec:cultural_evolution}

\subsection{Tool use}
\label{sec:tool-use}



\bibliographystyle{plainnat}
\bibliography{evo}


\makeindex

\end{document}
